---
title: "actfts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{actfts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(actfts)
```



# Motivation



# Data

This package features a three-time series that automatically updates from the FRED database of the United States. These datasets let you practice using the package's functions. Below, we describe each of them:

**Gross Domestic Product**, a comprehensive measure that quantifies the total monetary value of all goods and services produced within a country over a specific period, usually a quarter or a year. It provides a comprehensive overview of a nation's economic activity, reflecting its health and size. Economists often use GDP to compare the economic performance of different countries or regions and assess economic policies' impact.

Here's a practical code that allows you to obtain GDP data.

```{r}
GDP_data <- actfts::GDPEEUU
head(GDP_data)
```

**Personal Consumption Expenditures** represents the total value of goods and services that households and nonprofit institutions serving households consume within an economy. As a critical component of GDP, PCE reveals consumer behavior and spending patterns. It covers expenditures on durable goods, non-durable goods, and services, offering insights into consumer confidence and living standards.

Here's a practical code that allows you to obtain PCE data.

```{r}
PCEC_data <- actfts::PCECEEUU
head(PCEC_data)
```

**Disposable Personal Income** indicates how much money households spend and save after paying taxes and other mandatory charges. It serves as a crucial indicator of consumer purchasing power and financial well-being. DPI directly influences consumer spending and saving decisions, affecting overall economic growth. Analysts often examine DPI to understand trends in personal savings rates and consumption patterns.

Here's a practical code that allows you to obtain DPI data.

```{r}
DPI_data <- actfts::DPIEEUU
head(DPI_data)
```

# Applied Examples

```{r}
result <- acfinter(GDP_data, lag = 10)
print(result)
```

```{r}
result <- acfinter(GDP_data, lag = 10, ci.method = "ma", ci = 0.98)
print(result)
```

```{r}
result <- acfinter(GDP_data, lag = 10, delta = "diff1")
print(result)
```

# Final considerations

This tool can be used for time series analysis with an xts class condition; therefore, the user can transform the series to ts if any tool conflicts with an xts series.

# References

* U.S. Bureau of Economic Analysis, Gross Domestic Product (GDP), retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/GDP.
* U.S. Bureau of Economic Analysis, Personal Consumption Expenditures (PCEC), retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/PCEC
* U.S. Bureau of Economic Analysis, Disposable Personal Income (DPI), retrieved from FRED, Federal Reserve Bank of St. Louis;https://fred.stlouisfed.org/series/DPI




