---
title: "actfts"
output:
  rmarkdown::html_vignette:
    self_contained: false
vignette: >
  %\VignetteIndexEntry{actfts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
  body {
    text-align: justify;
  }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(actfts)
```

# Motivation

This package consolidates various R time series functions related to econometric and financial analysis, integrating these tools into a single package. Educators and researchers can understand time series analysis without the burden of fragmented code spread across multiple resources, creating coherent and efficient results reading.

The package simplifies and manages scattered code in an educational environment, allowing students to focus on data analysis and result interpretation. Moreover, gain a deeper and more practical understanding of the concepts without being overwhelmed by navigating multiple tools and syntaxes.

# Demo Data

This package includes a three-time series that automatically updates from the FRED database of the United States. These datasets allow you to practice using the package’s functions. Below is a brief description of each:

**Gross Domestic Product**, This measure quantifies the total monetary value of all goods and services produced within a country over a specific period, typically a quarter or a year. It provides a comprehensive overview of a nation’s economic activity, reflecting its size and economic health. Economists often use GDP to compare economic performance across countries or regions and assess the impact of economic policies.

Below is a practical code example to retrieve GDP data.

```{r}
GDP_data <- actfts::GDPEEUU
head(GDP_data)
```

**Personal Consumption Expenditures** This represents the total value of goods and services consumed by households and nonprofit institutions serving households within an economy. As a critical component of GDP, PCE reveals consumer behavior and spending patterns. It covers expenditures on durable goods, non durable goods, and services, providing insights into consumer confidence and living standards.

Below is a handy code that allows you to retrieve PCE data

```{r}
PCEC_data <- actfts::PCECEEUU
head(PCEC_data)
```

**Disposable Personal Income** refers to the amount of money households have available for spending and saving after deducting taxes and other mandatory charges. It is an indicator of consumer purchasing power and financial health. DPI significantly impacts consumer spending and saving behaviors, affecting overall economic growth. Analysts frequently study DPI to identify trends in personal savings rates and consumption patterns.

Below is a practical code snippet for obtaining DPI data.

```{r}
DPI_data <- actfts::DPIEEUU
head(DPI_data)
```

# Previews Concepts :

Before beginning with the examples, define the concepts that the "acfinter()" is the function used to calculate essentials

### ACF (Autocorrelation Function)

ACF measures the correlation between a time series and its lags at different intervals. It helps identify patterns of temporal dependence (Box & Jenkins, 1976).

### PACF (Partial Autocorrelation Function)

PACF measures the correlation between a time series and its lags, removing the influence of the intermediate lags. It is essential to determine the appropriate order in an AR(p) model (Box & Jenkins, 1976).

### Ljung-Box Test

The Ljung-Box Test statistically evaluates whether the autocorrelations of a time series differ from zero. Analysts commonly use it to check the independence of residuals in ARIMA models (Ljung & Box, 1978)

### Box-Pierce Test

Similar to the Ljung-Box test, the Box-Pierce Test is a statistical test that evaluates the independence of residuals in a time series model. It is a simpler and less refined version compared to the Ljung-Box test (Box & Pierce, 1970)

### KPSS Test

The KPSS Test (Kwiatkowski-Phillips-Schmidt-Shin) is a statistical test that checks the stationarity of a time series, mainly whether a series is stationary around its mean or a deterministic trend (Kwiatkowski et al., 1992)

### Box-Cox Transformation

The Box-Cox Transformation is a data transformation technique that applies a power function to stabilize variance and make the time series more closely resemble a normal distribution (Box & Cos, 1964)

### Kolmogorov-Smirnov Test

The Kolmogorov-Smirnov Test is a nonparametric test that compares a sample distribution with a reference probability distribution or compares two sample distributions to assess whether they differ significantly (Kolmogorov, 1933) (Smirnov, 1948)

### Shapiro-Wilk Test

The Shapiro-Wilk Test is a statistical test that evaluates the normality of a data set. It is particularly effective for small sample sizes and assesses whether a sample comes from a normally distributed population (Shapiro & Wilk, 1965)

# Applied Example

## Example one

In this first example, we will analyze the United States GDP time series. We will use the `acfinter()` function to see the ACF, PACF, Box_Pierce, Pv_Box, Ljung_Box, and Pv_Ljung. We can also see the normality analysis and stationarity analysis. Finally, a plot with ACF, PACF, and Pv LJUNG BOX.

Therefore, we will use the following code to analyze the first ten lags and explain each result of the `acfinter()` function.

```{r}
result <- acfinter(GDP_data, lag = 10)
print(result)
```

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, , fig.dim=c(7.6, 4)}
library(dplyr)
library(plotly)

ldata <- nrow(GDP_data)
result <- acfinter(GDP_data, lag = 10)
table <- result$`ACF-PACF Test`

get_clim1 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$acf^2)))
    return(clim[-length(clim)])
  } else {
    lineci1 <- rep(clim0, NROW(table$acf))
    return(lineci1)
  }
}

get_clim2 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$pacf^2)))
    return(clim[-length(clim)])
  } else {
    lineci2 <- rep(clim0, NROW(table$pacf))
    return(lineci2)
  }
}

saveci1 <- get_clim1(table$acf)
saveci2 <- get_clim2(table$pacf)

lag <- 10

fig1 <- plot_ly(
  x = table$lag,
  y = table$acf,
  type = "bar",
  name = "acf",
  color = I("slategray"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig1 <- fig1 %>% add_trace(
  y = saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig1 <- fig1 %>% add_trace(
  y = -saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- plot_ly(
  x = table$lag,
  y = table$pacf,
  type = "bar",
  name = "pacf",
  color = I("dimgrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig2 <- fig2 %>% add_trace(
  y = saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- fig2 %>% add_trace(
  y = -saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

hline <- function(y = 0, color = "black") {
  list(type = "line", x0 = 0, x1 = 1, xref = "paper", y0 = y, y1 = y,
       line = list(color = color, dash = "dash")
  )
}

fig3 <- plot_ly(
  x = table$lag,
  y = table$Pv_Ljung,
  type = "scatter",
  mode = "markers",
  name = "Pv Ljung Box",
  color = I("lightslategrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(shapes = hline(0.05),
         xaxis = list(range = c(0.5,lag+0.5)))

fig <- subplot(fig1, fig2, fig3, nrows = 3, shareX = TRUE, margin = 0.07) %>%
  layout(
    xaxis = list(
      title = "lags",
      dtick = 1,
      tick0 = 1,
      tickmode = "linear"
    )
  )

htmltools::tagList(fig)
```

The `result` variable contains three primary tests: ACF-PACF, Stationary, and Normality. For the ACF-PACF component, we observe significant correlations at multiple lags, and the Box Pierce and Ljung Box tests corroborate this finding.

Regarding the stationary tests, the `acfinter()` function provides ADF, KPSS-Level, KPSS-Trend, and PP results. While each test has nuances, they aim to determine if the series is stationary. Both the ADF and PP tests yield p-values of 0.99. Since these values exceed the significance level (in this case, 0.95), we don't reject the null hypothesis of stationarity, confirming the presence of a unit root and a trend component in the GDP time series.

On the other hand, we have the KPSS-Level and KPSS-Trend tests. Both the KPSS-Level and KPSS-Trend tests yielded p-values of 0.01. The result indicates that we reject the null hypothesis for both tests, suggesting that the time series is non-stationary, neither at level nor trend.

Concerning the third component, based on the Shapiro-Wilks and Kolmogorov-Smirnov normality tests, we concluded that the time series data does not follow a normal distribution. This is because the results led us to reject the null hypothesis of normality.

Finally, we conducted a Box-Cox transformation test to determine if a transformation was necessary to stabilize the time series variance. Based on the test results, we can apply some of these transformations:

* 1: No transformation.
* 0: Natural logarithm.
* 0.5: Square root.
* 2: Square.
* -1: Inverse.

The results in both tests are near 0, so the tests show we should apply to natural logarithm transformation.

## Example two

We've discussed the output of the `acfinter()` function. In the following examples, we'll explore how to use each function argument and provide important considerations for its correct application.

Let's examine the `ci.method` and `ci` arguments. The `ci.method` argument lets us select either constant confidence intervals ("white") or dynamic confidence intervals ("ma"). As for the `ci` argument, you can specify any desired value between 0 to 0.99.

If we can use these arguments, we will define to `ci.method` argument ma value whereas `ci` argument we will define a confidence interval of 0.98. So, we can use the following code below:

```{r}
result <- acfinter(GDP_data, lag = 10, ci.method = "ma", ci = 0.98)
print(result)
```

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, , fig.dim=c(7.6, 4)}
library(dplyr)
library(plotly)

ldata <- nrow(GDP_data)
result <- acfinter(GDP_data, lag = 10, ci.method = "ma", ci = 0.98)
table <- result$`ACF-PACF Test`

get_clim1 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$acf^2)))
    return(clim[-length(clim)])
  } else {
    lineci1 <- rep(clim0, NROW(table$acf))
    return(lineci1)
  }
}

get_clim2 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$pacf^2)))
    return(clim[-length(clim)])
  } else {
    lineci2 <- rep(clim0, NROW(table$pacf))
    return(lineci2)
  }
}

saveci1 <- get_clim1(table$acf,ci.type="ma", ci = 0.98)
saveci2 <- get_clim2(table$pacf,ci.type="ma", ci = 0.98)

lag <- 10

fig1 <- plot_ly(
  x = table$lag,
  y = table$acf,
  type = "bar",
  name = "acf",
  color = I("slategray"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig1 <- fig1 %>% add_trace(
  y = saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig1 <- fig1 %>% add_trace(
  y = -saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- plot_ly(
  x = table$lag,
  y = table$pacf,
  type = "bar",
  name = "pacf",
  color = I("dimgrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig2 <- fig2 %>% add_trace(
  y = saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- fig2 %>% add_trace(
  y = -saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

hline <- function(y = 0, color = "black") {
  list(type = "line", x0 = 0, x1 = 1, xref = "paper", y0 = y, y1 = y,
       line = list(color = color, dash = "dash")
  )
}

fig3 <- plot_ly(
  x = table$lag,
  y = table$Pv_Ljung,
  type = "scatter",
  mode = "markers",
  name = "Pv Ljung Box",
  color = I("lightslategrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(shapes = hline(0.05),
         xaxis = list(range = c(0.5,lag+0.5)))

fig <- subplot(fig1, fig2, fig3, nrows = 3, shareX = TRUE, margin = 0.07) %>%
  layout(
    xaxis = list(
      title = "lags",
      dtick = 1,
      tick0 = 1,
      tickmode = "linear"
    )
  )

htmltools::tagList(fig)
```

As you can see, the `acfinter()` function with the `ci.method` and `ci` arguments adjust the results based on the specified confidence interval. Additionally, we observe that the plot's confidence intervals are not constant but exhibit dynamic behavior.

## Example 3

To continue with the United States GPD analysis, we can use the delta argument to check the first three differences of the time series. We will analyze the first difference for this case and use the following code.

```{r}
result <- acfinter(GDP_data, lag = 10, delta = "diff1")
print(result)
```

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, , fig.dim=c(7.6, 4)}
library(dplyr)
library(plotly)

ldata <- nrow(GDP_data)
result <- acfinter(GDP_data, lag = 10, delta = "diff1")
table <- result$`ACF-PACF Test`

get_clim1 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$acf^2)))
    return(clim[-length(clim)])
  } else {
    lineci1 <- rep(clim0, NROW(table$acf))
    return(lineci1)
  }
}

get_clim2 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$pacf^2)))
    return(clim[-length(clim)])
  } else {
    lineci2 <- rep(clim0, NROW(table$pacf))
    return(lineci2)
  }
}

saveci1 <- get_clim1(table$acf)
saveci2 <- get_clim2(table$pacf)

lag <- 10

fig1 <- plot_ly(
  x = table$lag,
  y = table$acf,
  type = "bar",
  name = "acf",
  color = I("slategray"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig1 <- fig1 %>% add_trace(
  y = saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig1 <- fig1 %>% add_trace(
  y = -saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- plot_ly(
  x = table$lag,
  y = table$pacf,
  type = "bar",
  name = "pacf",
  color = I("dimgrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig2 <- fig2 %>% add_trace(
  y = saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- fig2 %>% add_trace(
  y = -saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

hline <- function(y = 0, color = "black") {
  list(type = "line", x0 = 0, x1 = 1, xref = "paper", y0 = y, y1 = y,
       line = list(color = color, dash = "dash")
  )
}

fig3 <- plot_ly(
  x = table$lag,
  y = table$Pv_Ljung,
  type = "scatter",
  mode = "markers",
  name = "Pv Ljung Box",
  color = I("lightslategrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(shapes = hline(0.05),
         xaxis = list(range = c(0.5,lag+0.5)))

fig <- subplot(fig1, fig2, fig3, nrows = 3, shareX = TRUE, margin = 0.07) %>%
  layout(
    xaxis = list(
      title = "lags",
      dtick = 1,
      tick0 = 1,
      tickmode = "linear"
    )
  )

htmltools::tagList(fig)
```


You can see when we use the argument `delta` with a value equal to "diff1" The `acfinter()` function shows all results with the first difference of the time series. So, we will explain the results.

First, the ACF-PACF test reveals significant autocorrelation at the 10th lag. The Box-Pierce and Ljung-Box tests confirm this finding, and the plot supports these results.

The stationary test shows that the AFD and PP test confirm that the first difference in the GDP time series is stationary, as we reject the null hypothesis. However, the KPSS level and trend show that the first difference in the GDP time series is not stationary.

On the other hand, the normality test shows that the first difference in the GDP time series is not normally distributed, as both tests reject the null hypothesis.

We must consider this: the Box-Cox test to stabilize the variance only applies to level time series, not the diffs of the time series.

## Example 4

Finally, the `acfinter()` function lets you view the result interactively. To do this, you need to use the argument interactive and type "acftable". You can dynamically view the result.

Additionally, you can download the results of a time series analysis by setting the download argument to TRUE. This action will generate an Excel file with the numeric results and a PNG image at 300 dpi resolution, which includes the ACF, PACF, and Pv Ljung-Box plots. The files will be saved in your documents folder. Here is an example of the code to use:

```{r echo=TRUE}
result <- acfinter(GDP_data, lag = 10, interactive = "acftable", download = TRUE)
```

# Final considerations

You can analyze time series in xts, ts, integer, and vector (numeric) formats. So, if you use a different format, the `acfinter()` function won't work. In this case, you must convert your data to any format we initially showed you.

# References

1. Box, G. E. P., & Cox, D. R. (1964). An Analysis of Transformations. Journal of the Royal Statistical Society: Series B (Methodological), 26(2), 211-243.
2. Box, G. E. P., & Jenkins, G. M. (1976). Time Series Analysis: Forecasting and Control. Revised Edition, Holden Day, San Francisco.
3. Box, G. E. P., & Pierce, D. A. (1970). Distribution of Residual Autocorrelations in Autoregressive-Integrated Moving Average Time Series Models. Journal of the American Statistical Association, 65(332), 1509-1526.
4. Glaz, R. (2023). reactable: Interactive Data Tables Based on 'React Table' (R package version 0.4.4) [Software]. Retrieved from https://CRAN.R-project.org/package=reactable
5. Hyndman, R. J., & Khandakar, Y. (2008). forecast: Forecasting Functions for Time Series and Linear Models (R package version 8.15) [Software]. Retrieved from https://CRAN.R-project.org/package=forecast
6. Kolmogorov, A. N. (1933). Sulla Determinazione Empirica di una Legge di Distribuzione. Giornale dell'Istituto Italiano degli Attuari, 4, 83-91.
7. Kwiatkowski, D., Phillips, P. C. B., Schmidt, P., & Shin, Y. (1992). Testing the Null Hypothesis of Stationarity against the Alternative of a Unit Root: How Sure Are We that Economic Time Series Have a Unit Root? Journal of Econometrics, 54(1-3), 159-178.
8. Ljung, G. M., & Box, G. E. P. (1978). On a Measure of Lack of Fit in Time Series Models. Biometrika, 65(2), 297-303.
9. R Core Team. (2023). stats: A package for statistical calculations in R (R package version 4.3.1) [Software]. Retrieved from https://stat.ethz.ch/R-manual/R-devel/library/stats/html/00Index.html
10. Ryan, J. A., & Ulrich, J. M. (2020). xts: eXtensible Time Series (R package version 0.12-0) [Software]. Retrieved from https://CRAN.R-project.org/package=xts
11. Shapiro, S. S., & Wilk, M. B. (1965). An Analysis of Variance Test for Normality (Complete Samples). Biometrika, 52(3/4), 591-611.
12. Sievert, C. (2020). plotly: Create Interactive Web Graphics via 'plotly.js' (R package version 4.9.2.1) [Software]. Retrieved from https://CRAN.R-project.org/package=plotly
13. Smirnov, N. V. (1948). Table for Estimating the Goodness of Fit of Empirical Distributions. The Annals of Mathematical Statistics, 19(2), 279-281.
14. Trapletti, A., & Hornik, K. (2020). tseries: Time Series Analysis and Computational Finance (R package version 0.10-48) [Software]. Retrieved from https://CRAN.R-project.org/package=tseries
15. U.S. Bureau of Economic Analysis, Disposable Personal Income (DPI). Retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/DPI
16. U.S. Bureau of Economic Analysis, Gross Domestic Product (GDP). Retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/GDP
17. U.S. Bureau of Economic Analysis, Personal Consumption Expenditures (PCEC). Retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/PCEC
18. Walker, A. (2023). openxlsx: Read, Write and Edit xlsx Files (R package version 4.2.5) [Software]. Retrieved from https://CRAN.R-project.org/package=openxlsx
