---
title: "actfts"
output:
  rmarkdown::html_vignette:
    self_contained: false
vignette: >
  %\VignetteIndexEntry{actfts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
  body {
    text-align: justify;
  }
</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(actfts)
```

# Motivation

The development of this package addresses the need to consolidate various concepts related to time series analysis, which current R packages approach separately. By integrating these tools into one package, educators can focus on teaching and understanding time series analysis without dealing with fragmented code across multiple resources. This integration provides a more coherent and efficient teaching experience by creating a unified and streamlined environment.

Moreover, this package aims to simplify the process for students by eliminating the complexities of managing scattered code. Students can then concentrate on analyzing data and interpreting results. By easing access to and application of time series analysis techniques, the package helps students gain a deeper, more practical understanding of the concepts without being overwhelmed by the need to navigate multiple tools and syntaxes. This approach enhances the overall quality of teaching and learning in time series analysis.

# Data

This package features a three-time series that automatically updates from the FRED database of the United States. These datasets let you practice using the package's functions. Below, we describe each of them:

**Gross Domestic Product**, GDP is a comprehensive measure that quantifies the total monetary value of all goods and services produced within a country over a specific period, usually a quarter or a year. It provides a comprehensive overview of a nation's economic activity, reflecting its health and size. Economists often use GDP to compare the economic performance of different countries or regions and assess economic policies' impact.

Here's a practical code that allows you to obtain GDP data.

```{r}
GDP_data <- actfts::GDPEEUU
head(GDP_data)
```

**Personal Consumption Expenditures** represents the total value of goods and services that households and nonprofit institutions serving households consume within an economy. As a critical component of GDP, PCE reveals consumer behavior and spending patterns. It covers expenditures on durable goods, non-durable goods, and services, offering insights into consumer confidence and living standards.

Here's a practical code that allows you to obtain PCE data.

```{r}
PCEC_data <- actfts::PCECEEUU
head(PCEC_data)
```

**Disposable Personal Income** indicates how much money households spend and save after paying taxes and other mandatory charges. It serves as a crucial indicator of consumer purchasing power and financial well-being. DPI directly influences consumer spending and saving decisions, affecting overall economic growth. Analysts often examine DPI to understand trends in personal savings rates and consumption patterns.

Here's a practical code that allows you to obtain DPI data.

```{r}
DPI_data <- actfts::DPIEEUU
head(DPI_data)
```

# Applied Example

If you want to use actfts, perform the package installation process using `install.package("actfts")` or through CRAN and load the `library(actfts)` library. Once you install the package, use the `acfinter()` function to obtain the time series analysis. Therefore, we are going to show you the different arguments that `acfinter()` has to analyze time series:

In this first example, you can use the `acfinter()` function to analyze the n lags (in this case, ten lags) of the time series, where you can see a table with five columns with values: lag, ACF, PACF, Box_Pierce, Pv_Box, Ljung_Box and Pv_Ljung, normality analysis, and stationarity analysis—finally, a plot with ACF, PACF, and Pv LJUNG BOX.

```{r}
result <- acfinter(GDP_data, lag = 10)
print(result)
```

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, , fig.dim=c(7.6, 4)}
library(dplyr)
library(plotly)

ldata <- nrow(GDP_data)
result <- acfinter(GDP_data, lag = 10)
table <- result$`ACF-PACF Test`

get_clim1 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$acf^2)))
    return(clim[-length(clim)])
  } else {
    lineci1 <- rep(clim0, NROW(table$acf))
    return(lineci1)
  }
}

get_clim2 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$pacf^2)))
    return(clim[-length(clim)])
  } else {
    lineci2 <- rep(clim0, NROW(table$pacf))
    return(lineci2)
  }
}

saveci1 <- get_clim1(table$acf)
saveci2 <- get_clim2(table$pacf)

lag <- 10

fig1 <- plot_ly(
  x = table$lag,
  y = table$acf,
  type = "bar",
  name = "acf",
  color = I("slategray"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig1 <- fig1 %>% add_trace(
  y = saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig1 <- fig1 %>% add_trace(
  y = -saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- plot_ly(
  x = table$lag,
  y = table$pacf,
  type = "bar",
  name = "pacf",
  color = I("dimgrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig2 <- fig2 %>% add_trace(
  y = saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- fig2 %>% add_trace(
  y = -saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

hline <- function(y = 0, color = "black") {
  list(type = "line", x0 = 0, x1 = 1, xref = "paper", y0 = y, y1 = y,
       line = list(color = color, dash = "dash")
  )
}

fig3 <- plot_ly(
  x = table$lag,
  y = table$Pv_Ljung,
  type = "scatter",
  mode = "markers",
  name = "Pv Ljung Box",
  color = I("lightslategrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(shapes = hline(0.05),
         xaxis = list(range = c(0.5,lag+0.5)))

fig <- subplot(fig1, fig2, fig3, nrows = 3, shareX = TRUE, margin = 0.07) %>%
  layout(
    xaxis = list(
      title = "lags",
      dtick = 1,
      tick0 = 1,
      tickmode = "linear"
    )
  )

htmltools::tagList(fig)
```

Now, the `acfinter()` function allows you to define the type and value of the confidence interval. It would help if you used the `ci.method` and `ci` argument; in the ci.method, you can choose constant confidence intervals with a "white" value or dynamic confidence intervals with a "ma" value. Concerning the ci argument, you can type any value. Here's an example:

```{r}
result <- acfinter(GDP_data, lag = 10, ci.method = "ma", ci = 0.98)
print(result)
```

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, , fig.dim=c(7.6, 4)}
library(dplyr)
library(plotly)

ldata <- nrow(GDP_data)
result <- acfinter(GDP_data, lag = 10, ci.method = "ma", ci = 0.98)
table <- result$`ACF-PACF Test`

get_clim1 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$acf^2)))
    return(clim[-length(clim)])
  } else {
    lineci1 <- rep(clim0, NROW(table$acf))
    return(lineci1)
  }
}

get_clim2 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$pacf^2)))
    return(clim[-length(clim)])
  } else {
    lineci2 <- rep(clim0, NROW(table$pacf))
    return(lineci2)
  }
}

saveci1 <- get_clim1(table$acf,ci.type="ma", ci = 0.98)
saveci2 <- get_clim2(table$pacf,ci.type="ma", ci = 0.98)

lag <- 10

fig1 <- plot_ly(
  x = table$lag,
  y = table$acf,
  type = "bar",
  name = "acf",
  color = I("slategray"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig1 <- fig1 %>% add_trace(
  y = saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig1 <- fig1 %>% add_trace(
  y = -saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- plot_ly(
  x = table$lag,
  y = table$pacf,
  type = "bar",
  name = "pacf",
  color = I("dimgrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig2 <- fig2 %>% add_trace(
  y = saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- fig2 %>% add_trace(
  y = -saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

hline <- function(y = 0, color = "black") {
  list(type = "line", x0 = 0, x1 = 1, xref = "paper", y0 = y, y1 = y,
       line = list(color = color, dash = "dash")
  )
}

fig3 <- plot_ly(
  x = table$lag,
  y = table$Pv_Ljung,
  type = "scatter",
  mode = "markers",
  name = "Pv Ljung Box",
  color = I("lightslategrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(shapes = hline(0.05),
         xaxis = list(range = c(0.5,lag+0.5)))

fig <- subplot(fig1, fig2, fig3, nrows = 3, shareX = TRUE, margin = 0.07) %>%
  layout(
    xaxis = list(
      title = "lags",
      dtick = 1,
      tick0 = 1,
      tickmode = "linear"
    )
  )

htmltools::tagList(fig)
```

You can analyze the time series at its level and at its differences. For this, you can use the delta argument to determine what kind of difference you would like to consider for analysis. Here's an example of the analysis of GDP at its first difference.

```{r}
result <- acfinter(GDP_data, lag = 10, delta = "diff1")
print(result)
```

```{r echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, , fig.dim=c(7.6, 4)}
library(dplyr)
library(plotly)

ldata <- nrow(GDP_data)
result <- acfinter(GDP_data, lag = 10, delta = "diff1")
table <- result$`ACF-PACF Test`

get_clim1 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$acf^2)))
    return(clim[-length(clim)])
  } else {
    lineci1 <- rep(clim0, NROW(table$acf))
    return(lineci1)
  }
}

get_clim2 <- function(x, ci=0.95, ci.type="white"){
  if (!ci.type %in% c("white", "ma")) stop('`ci.type` must be "white" or "ma"')
  clim0 <- qnorm((1 + ci)/2) / sqrt(ldata)
  if (ci.type == "ma") {
    clim <- clim0 * sqrt(cumsum(c(1, 2 * table$pacf^2)))
    return(clim[-length(clim)])
  } else {
    lineci2 <- rep(clim0, NROW(table$pacf))
    return(lineci2)
  }
}

saveci1 <- get_clim1(table$acf)
saveci2 <- get_clim2(table$pacf)

lag <- 10

fig1 <- plot_ly(
  x = table$lag,
  y = table$acf,
  type = "bar",
  name = "acf",
  color = I("slategray"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig1 <- fig1 %>% add_trace(
  y = saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig1 <- fig1 %>% add_trace(
  y = -saveci1,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- plot_ly(
  x = table$lag,
  y = table$pacf,
  type = "bar",
  name = "pacf",
  color = I("dimgrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(bargap = 0.7,
         xaxis = list(range = c(1,lag)))

fig2 <- fig2 %>% add_trace(
  y = saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

fig2 <- fig2 %>% add_trace(
  y = -saveci2,
  type = 'scatter',
  mode = 'lines',
  showlegend = FALSE,
  cliponaxis = FALSE,
  line = list(width = 0.8, dash = "dash", color="black")
)

hline <- function(y = 0, color = "black") {
  list(type = "line", x0 = 0, x1 = 1, xref = "paper", y0 = y, y1 = y,
       line = list(color = color, dash = "dash")
  )
}

fig3 <- plot_ly(
  x = table$lag,
  y = table$Pv_Ljung,
  type = "scatter",
  mode = "markers",
  name = "Pv Ljung Box",
  color = I("lightslategrey"),
  cliponaxis = FALSE,
  showlegend = FALSE
) %>%
  layout(shapes = hline(0.05),
         xaxis = list(range = c(0.5,lag+0.5)))

fig <- subplot(fig1, fig2, fig3, nrows = 3, shareX = TRUE, margin = 0.07) %>%
  layout(
    xaxis = list(
      title = "lags",
      dtick = 1,
      tick0 = 1,
      tickmode = "linear"
    )
  )

htmltools::tagList(fig)
```


Finally, the `acfinter()` function allows you to view the result interactively. To do this, you need to use the argument interactive and type "acftable" you can view the result in another way.

Additionally, you can download the results of a time series analysis. When you type "TRUE" in the download argument, the function downloads an Excel file with the numeric result and a figure in PNG format containing the ACF, PACF, and Pv Ljung Box Plots. Here's an example of the code you need to use it.

```{r eval=FALSE, include=FALSE}
result <- acfinter(GDP_data, lag = 10, interactive = "acftable", download = TRUE)
print(result)
```

# Final considerations

You can analyze time series in xts, ts, integer, and vector (numeric) formats. So, if you use a different format, the `acfinter()` function won't work. In this case, you must convert your data to any format we initially showed you.

# References

* U.S. Bureau of Economic Analysis, Gross Domestic Product (GDP), retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/GDP.
* U.S. Bureau of Economic Analysis, Personal Consumption Expenditures (PCEC), retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/PCEC
* U.S. Bureau of Economic Analysis, Disposable Personal Income (DPI), retrieved from FRED, Federal Reserve Bank of St. Louis;https://fred.stlouisfed.org/series/DPI




