---
title: "actfts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{actfts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning = FALSE, message = FALSE}
library(actfts)
```



# Motivation



# Data

**Gross Domestic Product**
```{r}
GDP_data <- actfts::GDPEEUU
head(GDP_data)
```

**Personal Consumption Expenditures (PCEC)**
```{r}
PCEC_data <- actfts::PCECEEUU
head(PCEC_data)
```

**Disposable Personal Income**
```{r}
DPI_data <- actfts::DPIEEUU
head(DPI_data)
```

# Applied Examples

```{r}
result <- acfinter(GDP_data, lag = 10)
print(result)
```

```{r}
result <- acfinter(GDP_data, lag = 10, ci.method = "ma", ci = 0.98)
print(result)
```

```{r}
result <- acfinter(GDP_data, lag = 10, delta = "diff1")
print(result)
```

# Final considerations

This tool can be used for time series analysis with an xts class condition; therefore, the user can transform the series to ts if any tool conflicts with an xts series.

# References

* U.S. Bureau of Economic Analysis, Gross Domestic Product (GDP), retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/GDP.
* U.S. Bureau of Economic Analysis, Personal Consumption Expenditures (PCEC), retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/PCEC
* U.S. Bureau of Economic Analysis, Disposable Personal Income (DPI), retrieved from FRED, Federal Reserve Bank of St. Louis;https://fred.stlouisfed.org/series/DPI




